---
title: "DATA 607 Week3 Assignment - Regular Expressions"
author: "Soumya Ghosh"
date: "September 13, 2018"
output: 
  html_document: 
    toc: true
    df_print: kable
    theme: cerulean
    highlight: pygments
    css: ./lab.css
  #  code_folding:hide
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### R Libraries:

Load necessary libraries -

```{r message=FALSE, warning=FALSE}
library(stringr)
```

### Problem 3:

Copy the introductory example. The vector name stores the extracted names.

```{r}
raw.data <- "555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson,Homer5553642Dr. Julius Hibbert"

name <- unlist(str_extract_all(raw.data, "[[:alpha:]., ]{2,}"))
name
```

### Part (a): Use the tools of this chapter to rearrange the vector so that all elements conform to the standard first_name last_name.:

1. There are some names where the name is abbreviate with initial followed by a '.' character. I want to replace them with BLANK in order to make them relatively easy to work with.


```{r}

name_1 <- str_replace_all(name,"\\s\\w{1}\\.", "") 
name_1 

```

2. There some names with titles. So in the ths step, I have removed the title section whereever applicable.

```{r}
name_2 <- str_replace_all(name_1, "\\w{2,3}\\.","")
name_2
```

3. Now the names separated by comma(,) need to be swapped in order to generate the First Name Last Name format.

```{r}
name_3 <- str_replace_all(name_2,"(\\w+),\\s(\\w+)","\\2 \\1")
name_3

```

### Part (b): Construct a logical vector indicating whether a character has a title (i.e., Rev. and Dr.):

```{r}
titlePresent <- str_detect(name,"^\\w+\\.")
titleResult <- data.frame(name, titlePresent)
titleResult

```


### Part (c): Construct a logical vector indicating whether a character has a second name:

```{r echo=TRUE}

secNamePresent <- str_detect(name,"\\s\\w{1}\\.")
secNameResult <- data.frame(name, secNamePresent)
secNameResult
```

### Problem 4:

Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.

### Part (a): [0-9]+\\$:

This expression represents one or more digits followed by a dollar symbol at the end. 

```{r echo=TRUE}

patternA <- "[0-9]+\\$"
sampleA <- c("12345$", "87564$","ahdghd$")
str_detect(sampleA,patternA)

```

### Part (b): \\b[a-z]{1,4}\\b:

This expression represents strings including 1 to 4 alphabets in lower case. 

```{r}

patternB <- "\\b[a-z]{1,4}\\b"
sampleB <- c("12345$", "abc","ahdgh")
str_detect(sampleB,patternB)

```

### Part (c): .*?\\.txt$:

This expression represents strings represents all file names ending with ".txt".

```{r}
patternC <- ".*?\\.txt$"
sampleC <- c("test123.txt", "test.xls","ahdgh")
str_detect(sampleC,patternC)

```

### Part (d): \\d{2}/\\d{2}/\\d{4}:

This expression represents a set of 2, 2 and 4 digits separated by a slash("/") character. Normally this format resents a date string in either MM/DD/YYYY or DD/MM/YYYY.


```{r}

patternD <- "\\d{2}/\\d{2}/\\d{4}"
sampleD <- c("01/20/2018", "28/02/2018","ahdgh")
str_detect(sampleD,patternD)

```

### Part (e): <(.+?)>.+?</\\1>:

This expression represents a HTML or XML tag in the format <Tag name>string in between</Tag Name>.

```{r}
patternE <- "<(.+?)>.+?</\\1>"
sampleE <- c("<a>This is a tag</a>", "28/02/2018","ahdgh")
str_detect(sampleE,patternE)
```

### Problem 9:

The following code hides a secret message. Crack it with R and regular expressions.
Hint: Some of the characters are more revealing than others! The code snippet is also
available in the materials at www.r-datacollection.com.

clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0Tanwo
Uwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigO
d6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5
fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr

```{r warning=FALSE}

cypherText <- "clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0TanwoUwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigO
d6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5
fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr"
```

**Step 1.** Replace all digits, lowercase characters and space/newline characters with BLANK.

```{r}
text1 <- str_replace_all(cypherText,"[[:digit:]|[:lower:]|[:space:]]+","")
text1

```

**Step 2.** Replace dot(.) with spaces.

```{r}

text2 <- str_replace_all(text1,"\\."," ")
text2

```

