---
title: "DATA 607 - Week1 Lab - Pittsburgh Bridges"
author: "Soumya Ghosh"
date: "September 1, 2018"
output:
  html_document: default
  pdf_document: default
---

## DATA 607 - Data Acquisition and Management - Week1 Hands on Lab
### Data Title: Pittsburgh Bridges Data Set 
### Source:

#### Creators: 

#### Yoram Reich & Steven J. Fenves 
#### Department of Civil Engineering and Engineering Design Research Center 
#### Carnegie Mellon University 
#### Pittsburgh, PA 15213 

#### Donor: 

#### Yoram Reich (yoram.reich '@' cs.cmu.edu)

#### The actual data set is found here: https://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1

### Load necessary R Libraries
```{r}
library(plyr)
library(htmlTable)
```


### Load bridges data into a Data Frame 

```{r}
bridges_df <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version1",header=FALSE,na.strings = "?")
head(bridges_df)

```

### Rename columns in the data frame

```{r}
names(bridges_df)[] <- c("V1"="Identifier","V2"= "River","V3"="Location","V4"="Erected","V5"="Purpose","V6"="Length","V7"="Lanes","V8"="Clear-G","V9"="T-Or-D","V10"="Material","V11"="Span","V12"="Rel-L","V13"="Type")
htmlTable(head(bridges_df))
```

### Summerize the bridges data

```{r}
summary(bridges_df)
```

### Distribution of bridges based on the year of erection/build

```{r}
hist(bridges_df$Erected)
```

### A pie chart for bridge purposes

```{r}
Purposes <- table(bridges_df$Purpose)
PurposeRatios <- Purposes/sum(Purposes)
PurposeLabels <- c("Aqueduct","Highway","Railroad","Walking")
pie(PurposeRatios, labels = PurposeLabels, main = "Purpose of Bridges Built")
```

### Railroad bridges - a histogram of the dates that they were installed:

```{r}

RRBridges <- subset(bridges_df, bridges_df$Purpose == "RR")
hist(RRBridges$Erected)
```

### Histogram of highway bridge dates of installation.

```{r}
HighwayBridges <- subset(bridges_df, bridges_df$Purpose == "HIGHWAY")
hist(HighwayBridges$Erected)
```