---
title: "Basic Sunburst"
author: "Jason Givens-Doyle"
date: "December 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries:

```{r}
library("sunburstR")
library("stringr")
library("tidyr")
```

Reading in the clean set first, will try getting the messy set next

```{r}
f1Sun<-read.csv("For Sunburst - Clean Fake Data.csv")  #this data is made up in my head, just wanted to get something out there
head(f1Sun)
```

```{r}
sunburst(f1Sun)
```


We encounter the small problem that cancer incidence, even in my fake data, is <1% of the population.  

```{r}
f1Sun2<-subset(f1Sun,str_detect(f1Sun$country,"Cancer")==TRUE)
f1Sun2$country<-str_replace(f1Sun2$country,"end","End")
sunburst(f1Sun2)
```

Using the "-end" is recomended for differentiation to avoid overlapping strings.  It is not needed here, but we can use it to make things look nice before getting rid of it.

```{r}
f1Sun3<-f1Sun2
f1Sun3$country<-str_replace(f1Sun3$country,"Cancer-End","Cancer-Other")
f1Sun3$country<-str_replace(f1Sun3$country,"-End","")
f1Sun3$country<-str_replace(f1Sun3$country,"-Cancer","")
sunburst(f1Sun3)
```

Now to do this with more normal data.  The idea is to get it mostly tidy, concatinate columns and then sunburst.

```{r}
f2Sun<-read.csv("For Sunburst - Fake Data.csv")
head(f2Sun)
```

I set this up with one column for incidence/mortality per 100,000 and another with the total numbers.  We won't have the total numbers, but that is trivial to fix if we have the population.  

```{r}
##First just ditching the columns we don't need
#
f2SunSmall<-data.frame(f2Sun$country,f2Sun$cancer.type,f2Sun$Total.Number.incidence,f2Sun$Total.number.mortality)
names(f2SunSmall)<-c("Country-Cancer","Type","Incidence","Mortality")
#
#Next pasting two of the columns together.  I don't have -end on the strings here, as above, it's not needed
#
f2SunSmall$`Country-Cancer`<-paste(f2SunSmall$`Country-Cancer`,f2SunSmall$Type,sep="-")
f2SunSmall<-data.frame(f2SunSmall$`Country-Cancer`,f2SunSmall$Incidence,f2SunSmall$Mortality)
names(f2SunSmall)<-c("Country-Cancer","Incidence","Mortality")
#
#Gathering the incidence and mortality numbers into a single column, then pasting the incidence/mortality indicator into the principle column
#
f2Sun2<-gather(f2SunSmall,`Incidence`,`Mortality`,key = "Diag",value="Patients")
f2Sun2$`Country-Cancer`<-paste(f2Sun2$`Country-Cancer`,f2Sun2$Diag,sep="-")
f2Sun2$Diag<-NULL
#deleting the word "Incidence", as we need to have a heirarchy
f2Sun2$`Country-Cancer`<-gsub("-Incidence","",f2Sun2$`Country-Cancer`)
```

##Discussion point

Here we need to make a decision on how to indicate mortality.  We can either show incidence then incidence for specific cancers and then mortality, or incidence then generally mortality then specific mortality.  I chose the first, but we could do the second.


```{r}
#need any - any - mortality, but not any - "All" - mortality, so subsetted that away
#
f2Sun3<-subset(f2Sun2,str_detect(f2Sun2$`Country-Cancer`,"-All-")==FALSE)
#
#Removing the other Alls from the strings
#
f2Sun3$`Country-Cancer`<-gsub("-All","",f2Sun3$`Country-Cancer`)
head(f2Sun3)
```

```{r}
#and now the graph
sunburst(f2Sun3,percent = TRUE)
sunburst(f2Sun3,percent = FALSE, count = TRUE)
```