---
title: "Movie Ratings Analysis"
author: "Soumya Ghosh"
date: "September 7, 2018"
output: 
  html_document: 
    toc: true
    df_print: kable
    theme: cerulean
    highlight: pygments
    css: ./lab.css
  #  code_folding:hide
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data Title: Movie Ratings Data Set 
### Source: Collected through indenpendent survey

### Database Components:

#### This analysis involves a MySQL source database called 'movieratings'. This database has 3 tables -
#### 1. movies
#### 2. users
#### 3. ratings

**Note:** MySQL database software need to be installed in the local computer in order to successfully execute the R markdown code.

#### Below are the steps involved in preparing the MySQL DB environment necesaary for the R markdown code to run successfully -

1. Run the DDL script (MovieRatings_DDL Script.sql) in order to create a database called 'movieratings' along with the necessary tables as mentioned above -
GitHub location:[DDL Script](https://github.com/soumya2g/R-CUNY-MSDS/blob/master/DATA-607/Movie%20Ratings/Source%20Files/MovieRatings_DDL%20Script.sql) 

2. Run the Data Load script (Data Load DML Script.sql) to load necessary data into the database -
GitHub Location: [Load Script](https://github.com/soumya2g/R-CUNY-MSDS/blob/master/DATA-607/Movie%20Ratings/Source%20Files/Data%20Load%20DML%20Script.sql)

**Note:** All the necessary CSV data files are also available in the below mention location -
GitHub Location: [Source Data Files](https://github.com/soumya2g/R-CUNY-MSDS/tree/master/DATA-607/Movie%20Ratings/Source%20Files)

3. Create a ODBC System DSN called 'MySQL_MovieRatings' in the local machine with 'movieratings' as the default database with appropriate root credentials or with an appropriate user id having READ access to 'movieratings' database.

### R Libraries:

Load necessary libraries -
```{r}
library(RODBC)
library(magrittr)
library(kableExtra)
library(data.table)
library(plyr)
library(ggplot2)
library(reshape2)
```

### Establish DB Connection:

I have used an ODBC data source called 'MySQL_MovieRatings' in order to connect to the database.

```{r}
con <- odbcConnect("MySQL_MovieRatings")
```

### Data Sets:

#### A. Movies:

I have taken a list of six Hollywood movies released in 2018 for this analysis. Movies belonging to multiple genres have been captured as pipe ('|') separated values.

```{r}
movies_df <- as.data.frame(sqlFetch(con,"movies"),stringsAsFactors = FALSE)

movies_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```

#### B. Users:

These users contributed in the movie rating survey process.

```{r}
users_df <- as.data.frame(sqlFetch(con,"users"),stringsAsFactors = FALSE)

users_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")
```

#### C. Ratings:

Each user rated all the 6 movies which has been loaded through SQL scripts into the database. A rank has been created in SQL based on the ratings provided by the users using SQL window function.

```{r}
Query = "SELECT 
             User_ID
            ,Movie_ID
            ,Rating
            ,RANK() OVER (PARTITION BY User_ID ORDER BY Rating desc) AS Movie_Rank
         FROM
             ratings 
         ORDER BY 
             User_ID
            ,Movie_Rank;"

ratings_df <- sqlQuery(con,Query)

ratings_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")
```

### Prepare Analytics data:

Here I have joined all the 3 data sets and loaded into a data frame to prepare for further analysis along with dynamically created movie rank informtaion.

```{r}
Query = "SELECT 
          m.Movie_ID
         ,m.Movie_Title
         ,m.Director_Name
         ,m.Movie_Year
         ,u.User_ID
         ,u.First_Name
         ,u.Last_Name
         ,u.Gender
         ,u.Age
         ,r.Rating
         ,RANK() OVER (PARTITION BY r.User_ID ORDER BY r.Rating desc) AS Movie_Rank
        FROM
          movieratings.ratings as r
        INNER JOIN
          movieratings.movies as m
        ON r.Movie_ID = m.Movie_ID
        INNER JOIN
        movieratings.users as u
        ON r.User_ID = u.User_ID;"

detail_ratings_df <- sqlQuery(con,Query)

detail_ratings_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```


### Data Transformtaion:

I have parsed the pipe('|') delimited genre column in the movies data frame to identify what all genres a particular movie belongs to as separate columns.

```{r echo=TRUE}
genres_df <- data.frame(sapply(tstrsplit(movies_df$Movie_Genre,'[|]',type.convert = TRUE),c),stringsAsFactors = FALSE)

genres_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```


I have Prepared a matrix with all the movie genres and assigned an indicator '1' if the genre applicable for a given movie.

```{r echo=TRUE}
genre_list <- c("Action","Adventure","Animation","Drama","Fantasy","Horror","Sci-Fi","Thriller")
genre_matrix <- matrix(0,nrow(genres_df)+1,length(genre_list))
genre_matrix[1,] <- genre_list #set first row to genre list
colnames(genre_matrix) <- genre_list

#iterate through matrix
for (i in 1:nrow(genres_df)) {
 for (c in 1:ncol(genres_df)) {
 genmat_col = which(genre_matrix[1,] == genres_df[i,c])
 genre_matrix[i+1,genmat_col] <- 1
 }
}

genre_matrix %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```


Converted the matrix into a data frame with integer indcator values.

```{r}
genre_matrix2 <- as.data.frame(genre_matrix[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list
for (c in 1:ncol(genre_matrix2)) {
  genre_matrix2[,c] <- as.integer(genre_matrix2[,c])
} #convert from characters to integers

genre_matrix2 %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")
```

I have combined the genre indicators to the movies_df data frame.

```{r}
movies_df <- cbind(movies_df,genre_matrix2)
movies_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```

Here I have joined the Data Frames 'detail_ratings_df' and 'movies_df' based on 'Movie_ID' to add genres related indicators to analytical data sets.

```{r}

detail_ratings_df <- merge(x = detail_ratings_df, y = subset(movies_df,select = c("Movie_ID",genre_list)), by = "Movie_ID")
detail_ratings_df %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

```

### Data Analysis:

Structure of the analytical data set.

```{r}
str(detail_ratings_df)
```

#### Movie Ratings Analysis:

Here I have created a side by side Boxplot for Ratings by Movie titles. This plot provides a comparative between the 6 selected titles. 

```{r warning=FALSE}
p<-ggplot(detail_ratings_df, aes(x = Movie_Title, y = Rating, color = Movie_Title)) +
  geom_boxplot()
p + theme(legend.position="bottom") + labs(title = "Movie Rating Boxplot By Title", x = "Movie Title") + theme(
plot.title = element_text(color="black", size=14, face="bold.italic"),
axis.title.x = element_text(color="blue", size=14, face="bold"),
axis.title.y = element_text(color="#993333", size=14, face="bold")
)
```

#### Movie Genre Analysis:

Here, I have calculate Mean Percentage of Movie Genre by Gender:

```{r warning=FALSE,message=FALSE}
summary_df <- subset(detail_ratings_df,select = c(8,12:19))
summary_df <- aggregate(summary_df,by = list(summary_df$Gender),FUN=mean, na.action=na.omit)
summary_df <- summary_df[-2] 
summary_df <- cbind(summary_df,gender=c(1:2))
summary_df <-summary_df[,2:length(summary_df)]
summary_df.long <- melt(summary_df,id.vars="gender")

ggplot(summary_df.long,aes(x=variable,y=value,fill=factor(gender)))+
  geom_bar(stat="identity",position="dodge")+
  scale_fill_discrete(name="Gender",
                      breaks=c(1, 2),
                      labels=c("Female", "Male")) + labs(title = "Movie Genre Popularity By Gender", x = "Genre", y = "Mean Percentage")
```

